---
title: "BIN381 Milestone 3 - Person A: Modelling Strategy and Implementation"
author: "Alexander Solomon - Group P1"
date: "`r Sys.Date()`"
output:
  html_document:
    df_print: paged
---

This notebook covers Task 1: modelling strategy and implementation. It starts from Milestone 2's validated data set and train/test split and does not redo cleaning, integration or feature engineering. Three candidate classifiers are built in a `tidymodels` workflow, tuned with stratified cross-validation on the training set, and handed off: predictions to Person B (Rudi, Task 2/3) and fitted workflows plus an input schema to Person C (Timothy, Task 4).

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = TRUE, message = FALSE)
# Paths below are relative to the project root; this lets the notebook knit from notebook/ too.
if (basename(getwd()) == "notebook") knitr::opts_knit$set(root.dir = normalizePath(".."))
library(tidyverse)
library(tidymodels)
library(rpart)
library(ranger)
library(glmnet)
library(knitr)

IN_DIR <- "data/integrated/"
OUT_DIR <- "data/modelling"
dir.create(OUT_DIR, showWarnings = FALSE, recursive = TRUE)

set.seed(381)
```

# 1.1 Analytical Question and Task Type

This is a multi-class classification problem: Business Objective 2 asks for households to be classified into food-insecurity vulnerability tiers. The target is `Vulnerability_Tier` (Food secure / Food insecure / Severely food insecure), derived in Milestone 2 from `FI_Score` on 20,925 households. `Food_Insecure_Binary` is kept as a fallback in case the thin middle tier (10.55% of households) is hard to separate, but the 3-tier target is primary.

# 1.2 Load the Milestone 2 Split

The train/test split is not re-created here. Milestone 2 already produced a stratified 70/30 split (`seed = 381`) on the validated, leakage-checked data set, so `M2_train.rds` and `M2_test.rds` are loaded as-is rather than re-split.

```{r load}
train_raw <- readRDS(file.path(IN_DIR, "M2_train.rds"))
test_raw  <- readRDS(file.path(IN_DIR, "M2_test.rds"))

cat("Train:", nrow(train_raw), " Test:", nrow(test_raw),
    " Total:", nrow(train_raw) + nrow(test_raw), "\n")
stopifnot(nrow(train_raw) + nrow(test_raw) == 20925)
```

## Leakage check

`train_raw` is Milestone 2's full integrated file (128 columns), so the raw food-insecurity survey items are still present at this point - that's expected. `leakage_vars` lists everything that must never reach the model; the actual check happens in 1.3, once predictors are selected.

```{r leakage-check}
leakage_vars <- c(
  "WorriedFoodWouldRunOutCode", "UnableToEatHealthyFoodCode", "AteFewFoodsCode",
  "SkippedMealCode", "AteLessCode", "RanOutOfFoodCode", "HungryButDidNotEatCode",
  "WholeDayWithoutEatingCode", "AdultHungerCode", "ChildHungerCode",
  "FewFoodsDays", "SkippedMealDays", "AteLessDays", "RanOutDays",
  "FI_Score", "n_fi_answered"
)

cat(ncol(train_raw), "columns loaded; leakage check runs once predictors are selected in 1.3.\n")
```

# 1.3 Predictor Set

The eight predictors are already engineered in Milestone 2 and reused as-is (`Income_Log` is `log1p()`-transformed income, `Income_PerCapita` is income divided by household size).

```{r predictor-set}
predictor_vars <- c("ComparativeIncomeCode", "GeoTypeCode", "HouseholdSize",
                    "Income_Log", "Income_PerCapita", "ElectricityAccessCode",
                    "SocialGrantRecipients", "MainToiletCode")
stopifnot(all(predictor_vars %in% names(train_raw)))
stopifnot(length(intersect(predictor_vars, leakage_vars)) == 0)  # predictor list is leakage-free

# Relabel tier levels to underscores so tidymodels' .pred_<level> columns don't contain spaces.
tier_lookup <- tibble(
  level = c("Food_secure", "Food_insecure", "Severely_food_insecure"),
  label = c("Food secure", "Food insecure", "Severely food insecure")
)

prep_df <- function(df) {
  df %>%
    mutate(
      Vulnerability_Tier = factor(Vulnerability_Tier, 
                                  levels = tier_lookup$label, 
                                  labels = tier_lookup$level),
      Food_Insecure_Binary = factor(Food_Insecure_Binary, 
                                    levels = c("Secure", "Insecure")),
      ComparativeIncomeCode = factor(ComparativeIncomeCode, 
                                     levels = 1:5, 
                                     ordered = TRUE),
      GeoTypeCode = factor(GeoTypeCode, 
                           levels = 1:3, 
                           labels = c("Urban", "Traditional_Tribal", "Farms")),
      ElectricityAccessCode = factor(ElectricityAccessCode, 
                                     levels = c(1, 2),
                                     labels = c("Yes", "No")),
      MainToiletCode = factor(MainToiletCode)
    ) %>%
    select(HouseholdID, Vulnerability_Tier, Food_Insecure_Binary, all_of(predictor_vars))
}

train <- prep_df(train_raw)
test  <- prep_df(test_raw)

present_leaks <- intersect(leakage_vars, names(train))
stopifnot(length(present_leaks) == 0)  # final predictor set is leakage-free

cat("Train class counts:\n"); print(table(train$Vulnerability_Tier))

cat("\nGeoTypeCode is 3 levels (Urban / Traditional-Tribal / Farms), corrected from the binary\n", 
    "urban/rural split Milestone 2's original encoding table assumed.\n")
```

# 1.4 Candidate Approaches and Assumptions

Three techniques, per the brief's requirement for at least two defensible candidates:

1.  **Regularised multinomial logistic regression** (`glmnet`) - the interpretable baseline. Assumes linearity in the log-odds; `Income_Log` already addresses the raw-income skew that would otherwise violate this. A VIF-style check below looks at redundancy between `ComparativeIncomeCode` and the two income features, since all three relate to household income.
2.  **Decision tree** (`rpart`) - gives directly readable rules for the interpretability discussion, at the cost of instability. Treated as an interpretability option, not the performance candidate.
3.  **Random forest** (`ranger`) - the performance benchmark. No scaling or dummy-encoding needed for tree-based models.

```{r multicollinearity-check}
vif_check <- train %>%
  mutate(CompIncomeNum = as.numeric(ComparativeIncomeCode)) %>%
  filter(!is.na(CompIncomeNum), !is.na(Income_Log), !is.na(Income_PerCapita))

m <- lm(CompIncomeNum ~ Income_Log + Income_PerCapita, data = vif_check)
r2 <- summary(m)$r.squared

cat("R-squared of ComparativeIncomeCode on the two income features:", 
    round(r2, 3), "\n")
cat("VIF-equivalent (1 / (1 - R2)):", 
    round(1 / (1 - r2), 2), "\n")
# VIF above ~5 is the usual threshold of concern
```

# 1.5 Recipes

Linear models need dummy-encoded, normalised predictors; tree-based models handle factors and unscaled numerics natively, so each gets its own recipe.

```{r recipes}
rec_linear <- recipe(Vulnerability_Tier ~ ., data = train %>% select(-HouseholdID, -Food_Insecure_Binary)) %>%
  step_impute_median(all_numeric_predictors()) %>%
  step_impute_mode(all_nominal_predictors()) %>%
  step_other(MainToiletCode, threshold = 0.02) %>%
  step_novel(all_nominal_predictors()) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors()) %>%
  step_normalize(all_numeric_predictors())

rec_tree <- recipe(Vulnerability_Tier ~ ., data = train %>% select(-HouseholdID, -Food_Insecure_Binary)) %>%
  step_impute_median(all_numeric_predictors()) %>%
  step_impute_mode(all_nominal_predictors()) %>%
  step_other(MainToiletCode, threshold = 0.02) %>%
  step_novel(all_nominal_predictors())
```

# 1.6 Model Specifications

```{r specs}
spec_logistic <- multinom_reg(penalty = tune(), mixture = tune()) %>%
  set_engine("glmnet") %>%
  set_mode("classification")

spec_tree <- decision_tree(cost_complexity = tune(), tree_depth = tune(), min_n = tune()) %>%
  set_engine("rpart") %>%
  set_mode("classification")

spec_forest <- rand_forest(mtry = tune(), min_n = tune(), trees = 500) %>%
  set_engine("ranger", importance = "permutation") %>%
  set_mode("classification")

wf_logistic <- workflow() %>% add_recipe(rec_linear) %>% add_model(spec_logistic)
wf_tree     <- workflow() %>% add_recipe(rec_tree)   %>% add_model(spec_tree)
wf_forest   <- workflow() %>% add_recipe(rec_tree)   %>% add_model(spec_forest)
```

# 1.7 Stratified Cross-Validation and Class Imbalance

Stratified k-fold CV keeps each fold's class mix close to the training set's 69.45/10.55/20.00 split. Class weighting is compared against plain stratification below to assess imbalance directly, rather than just noting it.

```{r cv}
set.seed(381)
folds <- vfold_cv(train, v = 5, strata = Vulnerability_Tier)
print(folds)
```

```{r class-weight-comparison}
# Frequency-inverse class weights, for comparison against unweighted stratified CV.
class_weights <- train %>%
  count(Vulnerability_Tier) %>%
  mutate(weight = sum(n) / (n_distinct(Vulnerability_Tier) * n))
print(class_weights)
```

# 1.8 Tuning

Tuning grids use space-filling designs (10-15 candidates per model) to keep the 5-fold search within reasonable runtime.

```{r tuning-grids}
grid_logistic <- grid_space_filling(penalty(), mixture(), size = 12)
grid_tree <- grid_space_filling(cost_complexity(), tree_depth(), min_n(), size = 12)
grid_forest <- grid_space_filling(mtry(range = c(2, 8)), min_n(), size = 10)

metrics_tune <- metric_set(roc_auc, accuracy, bal_accuracy)
```

```{r tune-logistic}
set.seed(381)
tuned_logistic <- tune_grid(
  wf_logistic, resamples = folds, grid = grid_logistic,
  metrics = metrics_tune, control = control_grid(save_pred = TRUE)
)
show_best(tuned_logistic, metric = "roc_auc")
```

```{r tune-tree}
set.seed(381)
tuned_tree <- tune_grid(
  wf_tree, resamples = folds, grid = grid_tree,
  metrics = metrics_tune, control = control_grid(save_pred = TRUE)
)
show_best(tuned_tree, metric = "roc_auc")
```

```{r tune-forest}
set.seed(381)
tuned_forest <- tune_grid(
  wf_forest, resamples = folds, grid = grid_forest,
  metrics = metrics_tune, control = control_grid(save_pred = TRUE)
)
show_best(tuned_forest, metric = "roc_auc")
```

Hyper parameters are selected on macro-averaged `roc_auc`, agreed as the Day-1 placeholder in the work split. Accuracy is reported alongside since the 69.45% majority class makes accuracy alone misleading. Person B's Task 2 makes the final metric call for model selection; if that differs, only the `select_best()` calls below need to change.

# 1.9 Finalise and Fit on Full Training Set

```{r finalize}
best_logistic <- select_best(tuned_logistic, metric = "roc_auc")
best_tree <- select_best(tuned_tree,     metric = "roc_auc")
best_forest <- select_best(tuned_forest,   metric = "roc_auc")

final_logistic <- finalize_workflow(wf_logistic, best_logistic) %>% fit(data = train)
final_tree <- finalize_workflow(wf_tree, best_tree) %>% fit(data = train)
final_forest <- finalize_workflow(wf_forest, best_forest) %>% fit(data = train)
```

# 1.10 Model Parameter Table

```{r parameter-table}
param_table <- tibble(
  Technique          = c("Regularised multinomial logistic regression", "Decision tree", "Random forest"),
  Engine = c("glmnet", "rpart", "ranger"),
  Final_Hyperparams = c(sprintf("penalty = %.4g, mixture = %.3f", 
                                best_logistic$penalty, best_logistic$mixture), 
                        sprintf("cost_complexity = %.4g, tree_depth = %d, 
                                min_n = %d", 
                                best_tree$cost_complexity, 
                                best_tree$tree_depth, best_tree$min_n), 
                        sprintf("mtry = %d, min_n = %d, trees = 500", 
                                best_forest$mtry, best_forest$min_n)),
  Resampling = "5-fold CV, stratified on Vulnerability_Tier",
  Seed = 381,
  Optimised_For = "roc_auc (macro, provisional - confirm with Person B)"
)
kable(param_table)
write_csv(param_table, file.path(OUT_DIR, "model_parameter_table.csv"))
```

# 1.11 Predictions Hand-off (Person B)

Both held-out test predictions and cross-validated training-fold predictions are exported, so Person B can compare CV performance against test performance for over fitting.

```{r predictions-handoff}
test_preds <- bind_rows(
  augment(final_logistic, test) %>% mutate(model = "logistic"),
  augment(final_tree, test) %>% mutate(model = "tree"),
  augment(final_forest, test) %>% mutate(model = "forest")
) %>%
  select(HouseholdID, model, Vulnerability_Tier, starts_with(".pred"))

cv_preds <- bind_rows(
  collect_predictions(tuned_logistic, parameters = best_logistic) %>% 
    mutate(model = "logistic"),
  collect_predictions(tuned_tree, parameters = best_tree) %>% 
    mutate(model = "tree"),
  collect_predictions(tuned_forest, parameters = best_forest) %>% 
    mutate(model = "forest")
)

write_csv(test_preds, file.path(OUT_DIR, "test_predictions_for_personB.csv"))
write_csv(cv_preds, file.path(OUT_DIR, "cv_predictions_for_personB.csv"))
write_csv(tier_lookup, file.path(OUT_DIR, "tier_label_lookup.csv"))

print(list.files(OUT_DIR, pattern = "personB|lookup"))
```

# 1.12 Model and Schema Hand-off (Person C)

Person C's Shiny app needs the fitted models and a schema of what a valid input row looks like, so input validation can be built without guessing column names, types or ranges.

```{r workflows-handoff}
saveRDS(final_logistic, file.path(OUT_DIR, "final_workflow_logistic.rds"))
saveRDS(final_tree,     file.path(OUT_DIR, "final_workflow_tree.rds"))
saveRDS(final_forest,   file.path(OUT_DIR, "final_workflow_forest.rds"))

input_schema <- tibble(
  variable = predictor_vars,
  r_type = c("ordered factor (1-5)", "factor (Urban/Traditional_Tribal/Farms)", 
             "numeric", "numeric (log1p of income)", "numeric (Rand per person)", 
             "factor (Yes/No)", "numeric (count)", "factor (1-12)"),
  valid_range = c("1 to 5; 9 = Unspecified -> reject/flag", "1, 2 or 3 only", 
                  "1 to 25; 0 is invalid -> reject; 8 and 9 ARE valid, do not reject", 
                  "derived from income, not user-entered directly", 
                  "derived from income and household size, not user-entered directly", 
                  "1 or 2; 9 = Unspecified -> reject/flag", 
                  "0 upward; 99 = Unspecified -> reject/flag", 
                  "1 to 12; 99 = Unspecified -> reject/flag"),
  missing_action = "flag and either impute (median/mode, matching the recipes above) or reject per Person C's validation design"
)
write_csv(input_schema, file.path(OUT_DIR, "input_schema_for_personC.csv"))
kable(input_schema)
```

# 1.13 Binary Fallback Model (App Interface Fix)

Person C's app (`app/R/score.R`) was built against a binary classifier with a single `.pred_insecure` probability column, before the 3-tier target was finalised as primary. Rather than wait on an app rewrite, a second model is fitted here on `Food_Insecure_Binary` - already in the data set as the documented fallback target - with its factor levels relabelled so the predicted column lands on exactly `.pred_insecure`. The app can use this unchanged. The 3-tier models above remain primary for the report and for Person B's evaluation.

```{r binary-prep}
binary_lookup <- tibble(level = c("secure", "insecure"), 
                        label = c("Secure", "Insecure")) 

train_binary <- train %>%   
  mutate(Food_Insecure_Binary = factor(Food_Insecure_Binary,
                                       levels = binary_lookup$label, 
                                       labels = binary_lookup$level)) 

test_binary <- test %>%   
  mutate(Food_Insecure_Binary = factor(Food_Insecure_Binary,
                                       levels = binary_lookup$label, 
                                       labels = binary_lookup$level))
```

```{r binary-recipe-spec}
rec_binary <- recipe(Food_Insecure_Binary ~ .,
                     data = train_binary %>% select(-HouseholdID, -Vulnerability_Tier)) %>%
  step_impute_median(all_numeric_predictors()) %>%
  step_impute_mode(all_nominal_predictors()) %>%
  step_other(MainToiletCode, threshold = 0.02) %>%
  step_novel(all_nominal_predictors()) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors()) %>%
  step_normalize(all_numeric_predictors())

spec_binary <- logistic_reg(penalty = tune(), mixture = tune()) %>%
  set_engine("glmnet") %>%
  set_mode("classification")

wf_binary <- workflow() %>% add_recipe(rec_binary) %>% add_model(spec_binary)
```

```{r binary-tune-finalize}
set.seed(381)
folds_binary <- vfold_cv(train_binary, v = 5, strata = Food_Insecure_Binary)
grid_binary <- grid_space_filling(penalty(), mixture(), size = 8)

tuned_binary <- tune_grid(
  wf_binary, resamples = folds_binary, grid = grid_binary,
  metrics = metric_set(roc_auc, accuracy, bal_accuracy),
  control = control_grid(save_pred = TRUE)
)
show_best(tuned_binary, metric = "roc_auc")

best_binary <- select_best(tuned_binary, metric = "roc_auc")
final_binary <- finalize_workflow(wf_binary, best_binary) %>% fit(data = train_binary)
```

```{r binary-export}
pred_check <- predict(final_binary, new_data = head(test_binary, 1), type = "prob")
stopifnot(".pred_insecure" %in% names(pred_check))

saveRDS(final_binary, file.path(OUT_DIR, "final_workflow_binary.rds"))
cat("Binary fallback model exported. Columns:", paste(names(pred_check), collapse = ", "), "\n")
```


# 1.14 Class Weights Compared with Plain Stratification

The work split asks for class imbalance to be handled with class weights or stratification, and for the two to be compared. Section 1.7 computed frequency-inverse class weights; here they are applied. The tuned multinomial logistic regression from 1.9 is refitted on the same five folds twice: once with stratified folds only (the setting used for every model above) and once with the class weights added as case weights. Nothing else differs.

```{r class-weight-fit}
train_w <- train %>%
  mutate(case_wt = hardhat::importance_weights(
    class_weights$weight[match(Vulnerability_Tier, class_weights$Vulnerability_Tier)]))

# Same steps as rec_linear; the importance-weight column gets the case-weights role automatically.
rec_linear_w <- recipe(Vulnerability_Tier ~ ., data = train_w %>% select(-HouseholdID, -Food_Insecure_Binary)) %>%
  step_impute_median(all_numeric_predictors()) %>%
  step_impute_mode(all_nominal_predictors()) %>%
  step_other(MainToiletCode, threshold = 0.02) %>%
  step_novel(all_nominal_predictors()) %>%
  step_dummy(all_nominal_predictors()) %>%
  step_zv(all_predictors()) %>%
  step_normalize(all_numeric_predictors())

spec_logistic_best <- finalize_model(spec_logistic, best_logistic)

set.seed(381)
folds_w <- vfold_cv(train_w, v = 5, strata = Vulnerability_Tier)
stopifnot(all(map2_lgl(folds$splits, folds_w$splits, ~ identical(.x$in_id, .y$in_id))))  # identical folds

metrics_cw <- metric_set(roc_auc, bal_accuracy, accuracy, brier_class)
ctrl_cw <- control_resamples(save_pred = TRUE)

res_plain <- fit_resamples(
  workflow() %>% add_recipe(rec_linear) %>% add_model(spec_logistic_best),
  resamples = folds, metrics = metrics_cw, control = ctrl_cw)
res_weighted <- fit_resamples(
  workflow() %>% add_recipe(rec_linear_w) %>% add_model(spec_logistic_best) %>% add_case_weights(case_wt),
  resamples = folds_w, metrics = metrics_cw, control = ctrl_cw)
```

```{r class-weight-compare}
summarise_cv <- function(res, label) {
  m <- collect_metrics(res) %>% select(.metric, mean) %>%
    pivot_wider(names_from = .metric, values_from = mean)
  r <- collect_predictions(res) %>%
    group_by(Vulnerability_Tier) %>%
    summarise(recall = mean(.pred_class == Vulnerability_Tier), .groups = "drop") %>%
    pivot_wider(names_from = Vulnerability_Tier, values_from = recall, names_prefix = "recall_")
  bind_cols(tibble(setting = label), m, r)
}
class_weight_cmp <- bind_rows(
  summarise_cv(res_plain,    "Stratified folds only"),
  summarise_cv(res_weighted, "Stratified folds + class weights")
) %>%
  select(setting, roc_auc, bal_accuracy, accuracy, brier_class,
         recall_Food_secure, recall_Food_insecure, recall_Severely_food_insecure)

kable(class_weight_cmp, digits = 3,
      caption = "Multinomial logistic regression, 5-fold CV: with and without class weights")
write_csv(class_weight_cmp, file.path(OUT_DIR, "class_weight_comparison.csv"))

pct <- function(x) sprintf("%.1f%%", 100 * x)
cw <- function(col, row) class_weight_cmp[[col]][row]
```

Class weights move the predictions towards the minority tiers. Recall for the middle tier rises from `r pct(cw("recall_Food_insecure", 1))` to `r pct(cw("recall_Food_insecure", 2))` and for the severe tier from `r pct(cw("recall_Severely_food_insecure", 1))` to `r pct(cw("recall_Severely_food_insecure", 2))`, but recall for food-secure households falls from `r pct(cw("recall_Food_secure", 1))` to `r pct(cw("recall_Food_secure", 2))` and accuracy from `r pct(cw("accuracy", 1))` to `r pct(cw("accuracy", 2))`. Ranking improves only slightly (Hand-Till AUC `r sprintf("%.3f", cw("roc_auc", 1))` against `r sprintf("%.3f", cw("roc_auc", 2))`), while the probabilities become much less trustworthy: the Brier score worsens from `r sprintf("%.3f", cw("brier_class", 1))` to `r sprintf("%.3f", cw("brier_class", 2))`, because the weights distort the base rates the model learns. Even with weights, about three in four middle-tier households are still missed, which supports the binary fallback in 1.13.

The unweighted models are therefore kept. Imbalance is handled by stratified folds and, for deployment, by a recall-based threshold on calibrated probabilities (Person B, Task 2). That reaches the recall target without distorting the probabilities the app shows to officials.
