---
title: "BIN381 Milestone 2 - Person A: Prepare for Modelling"
author: "Alexander Solomon - Group P1"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    df_print: paged
editor_options: 
  markdown: 
    wrap: 72
---

This notebook starts where `BIN 381 Milestone 2 Person B - Integration, Construction and Data Dictionary.Rmd` stops.
It reads the integrated household dataset from `data/integrated/`,
constructs and validates the Milestone 3 target variable, removes
leakage risks, finalises the predictor set from Milestone 1's hypotheses
(H1-H6), and produces a stratified train/test split. Everything it
produces goes back to `data/integrated/`.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = TRUE, message = FALSE)
# Paths are relative to the project root; this lets the notebook knit from notebook/ too.
if (basename(getwd()) == "notebook") knitr::opts_knit$set(root.dir = normalizePath(".."))
library(tidyverse)
library(rsample)
library(knitr)

IN_DIR  <- "data/integrated/"
OUT_DIR <- "data/integrated/"
dir.create(OUT_DIR, showWarnings = FALSE, recursive = TRUE)
```

## Load the integrated dataset

```{r load}
integrated <- readRDS(file.path(IN_DIR, "household_integrated.rds"))
stopifnot(nrow(integrated) == 20940)
cat("Households:", nrow(integrated), " Columns:", ncol(integrated), "\n")
```

## 4.1 Target Variable Construction

The Food Insecurity (FI) Score is the count of "Yes" (code 1) responses
across eight binary items in D07: `WorriedFoodWouldRunOutCode`,
`UnableToEatHealthyFoodCode`, `AteFewFoodsCode`, `SkippedMealCode`,
`AteLessCode`, `RanOutOfFoodCode`, `HungryButDidNotEatCode` and
`WholeDayWithoutEatingCode`.

The Milestone 1 EDA scored the raw D07 file directly and treated
sentinel code 9 (don't know/refused) as "No". Person C's Task 2 cleaning
made a more defensible choice for this integrated dataset: code 9 on
these items is recoded to a genuine `NA` rather than folded into "No",
since "don't know/refused" is not evidence of food security. That means
`FI_Score` here must be computed with missingness in mind, not just
summed blindly.

`AdultHungerCode` and `ChildHungerCode` are deliberately excluded from
the score: they use a different frequency scale (codes 1-6) rather than
binary yes/no, and mixing scales would distort the count. They are close
proxies for the same construct, so they are treated as leakage risks in
Section 4.4 below.

```{r fi-items}
fi_items <- c(
  "WorriedFoodWouldRunOutCode", "UnableToEatHealthyFoodCode",
  "AteFewFoodsCode", "SkippedMealCode", "AteLessCode",
  "RanOutOfFoodCode", "HungryButDidNotEatCode", "WholeDayWithoutEatingCode"
)
stopifnot(all(fi_items %in% names(integrated)))
```

## 4.1a Data Quality Finding: Households With No Valid FI Answers (DQ-11)

Because sentinel 9 is now `NA` rather than "No", a household can have a
partially or fully missing FI Score -- something the raw-file version of
this scoring never surfaced. Checking the integrated dataset directly,
before any scoring is done:

```{r dq11-check}
na_summary <- integrated %>%
  mutate(n_missing_items = rowSums(is.na(across(all_of(fi_items))))) %>%
  count(n_missing_items) %>%
  arrange(n_missing_items)

kable(na_summary, col.names = c("Missing FI items", "Households"))

n_any_missing <- integrated %>%
  mutate(n_missing_items = rowSums(is.na(across(all_of(fi_items))))) %>%
  filter(n_missing_items > 0) %>% nrow()
n_all_missing <- integrated %>%
  mutate(n_missing_items = rowSums(is.na(across(all_of(fi_items))))) %>%
  filter(n_missing_items == 8) %>% nrow()

cat("Households missing >=1 FI item:", n_any_missing, "\n")
cat("Households missing all 8 FI items (unscorable):", n_all_missing, "\n")
```

**Problem:** 15 households answered none of the 8 FI items, so no FI
Score can be computed for them. A set-intersection check against the 15
HouseholdIDs Person B repaired in Task 3 confirms this is a *different*
set of 15 households -- a distinct, previously undocumented data-quality
issue in the FI items themselves, not a symptom of the key-repair.

**Treatment:** `FI_Score` is set to `NA` when zero items are answered,
rather than defaulted to 0, which would silently and incorrectly record
these households as food secure. They are then explicitly excluded from
target construction and model training -- never imputed or guessed.

**Effect:** the modelling population becomes 20,925 of the 20,940
integrated households (99.93% retained).

```{r fi-score}
integrated <- integrated %>%
  mutate(
    n_fi_answered = rowSums(!is.na(across(all_of(fi_items)))),
    FI_Score      = if_else(n_fi_answered > 0,
                            rowSums(across(all_of(fi_items)) == 1, na.rm = TRUE),
                            NA_real_)
  )

n_unscorable <- sum(is.na(integrated$FI_Score))
cat("Excluded as unscorable:", n_unscorable, "\n")

modelling_data <- integrated %>% filter(!is.na(FI_Score))
cat("Retained for target construction:", nrow(modelling_data), "\n")
stopifnot(nrow(modelling_data) == 20940 - n_unscorable)
```

**Verification:** the FI Score distribution on the 20,925 retained
households, compared against the Milestone 1 raw-file report:

```{r fi-score-dist}
kable(table(FI_Score = modelling_data$FI_Score), col.names = c("FI_Score", "Households"))
```

## 4.2 Target Framing Decision

Two candidate framings are derived, both on the 20,925 households with a
valid FI Score:

```{r target-framing}
modelling_data <- modelling_data %>%
  mutate(
    Vulnerability_Tier = case_when(
      FI_Score == 0                  ~ "Food secure",
      FI_Score >= 1 & FI_Score <= 3  ~ "Food insecure",
      FI_Score >= 4                  ~ "Severely food insecure"
    ) %>% factor(levels = c("Food secure", "Food insecure", "Severely food insecure")),
    Food_Insecure_Binary = if_else(FI_Score >= 1, "Insecure", "Secure") %>%
      factor(levels = c("Secure", "Insecure"))
  )

cat("3-tier distribution (%):\n")
print(round(prop.table(table(modelling_data$Vulnerability_Tier)) * 100, 2))

cat("\nBinary distribution (%):\n")
print(round(prop.table(table(modelling_data$Food_Insecure_Binary)) * 100, 2))
```

**Decision: the 3-tier framing (Secure / Insecure / Severely insecure)
is the primary modelling target for Milestone 3.** This matches Business
Objective 2's own wording -- classifying households into "vulnerability
tiers" -- more directly than a binary split would, and it lets the
dashboard in Objective 3 distinguish moderate from severe cases.

**Documented contingency:** the middle tier (Insecure, \~10.5%) is
noticeably thinner than the other two. If Milestone 3 modelling shows
the 3-tier classifier cannot reliably separate the Insecure tier from
its neighbours, the team will fall back to the binary framing (Secure
vs. Insecure) as a documented, justified simplification rather than
silently changing the target.

## 4.3 Class Imbalance Assessment

Both framings are moderately imbalanced (majority class \~69.5% either
way). This is addressed at the train/test split stage (Section 4.6) via
stratified sampling rather than plain random sampling. If Milestone 3
model performance on the minority tiers is poor, class weighting or
resampling (e.g. SMOTE) should be considered then.

## 4.4 Leakage Variables Removed

The following are excluded from the predictor set because they define,
or closely proxy, the target itself -- cross-checked against Person B's
data dictionary, which flags the same 8 + 6 columns independently:

```{r leakage-list}
leakage_vars <- c(fi_items,
                  "AdultHungerCode", "ChildHungerCode",
                  "FewFoodsDays", "SkippedMealDays", "AteLessDays", "RanOutDays",
                  "FI_Score", "n_fi_answered")
stopifnot(all(leakage_vars[leakage_vars != "FI_Score" & leakage_vars != "n_fi_answered"] %in% names(integrated)))
print(leakage_vars)
```

`FI_Score` and `n_fi_answered` are included in this list once
`Vulnerability_Tier` / `Food_Insecure_Binary` are derived from them: a
model could otherwise infer the target from how many FI items a
household answered, not just what they answered.

## 4.5 Predictor Set & Encoding Plan

The predictor set is built around the six hypotheses (H1-H6) from
Milestone 1, so every retained feature has a stated analytical reason.
Income is cleaned and log-transformed per Milestone 1's DQ-09 finding
(right-skewed, sentinel 9999999 already converted to `NA` by Person C's
Task 2 cleaning):

```{r predictors}
predictor_vars <- c("ComparativeIncomeCode", "GeoTypeCode", "HouseholdSize",
                    "ElectricityAccessCode", "SocialGrantRecipients", "MainToiletCode")
stopifnot(all(predictor_vars %in% names(modelling_data)))
stopifnot("TotalMonthlyHouseholdIncomeRaw" %in% names(modelling_data))

modelling_data <- modelling_data %>%
  mutate(
    Income_Log       = log1p(TotalMonthlyHouseholdIncomeRaw),
    Income_PerCapita = TotalMonthlyHouseholdIncomeRaw / HouseholdSize
  )

cat("Income missingness:", sum(is.na(modelling_data$TotalMonthlyHouseholdIncomeRaw)),
    "of", nrow(modelling_data),
    sprintf("(%.2f%%)\n", 100 * mean(is.na(modelling_data$TotalMonthlyHouseholdIncomeRaw))))

summary(modelling_data$Income_Log)
```

| Variable | From | Encoding decision | Notes |
|------------------|------------------|------------------|------------------|
| ComparativeIncomeCode | H1 | Ordered factor (as surveyed) | Highest Cramer's V (0.28) of all categorical predictors in Milestone 1. |
| GeoTypeCode | H2 | Binary factor (urban/rural) | Test for interaction with income per H2's framing. |
| HouseholdSize | H3 | Numeric; also used to derive per-capita income | Per-capita income = income / HouseholdSize. |
| TotalMonthlyHouseholdIncomeRaw | H1, H3 | Cleaned (NA already handled by Person C) + log-transformed | Median vs mean confirms right skew (DQ-09); log1p() handles zero income. |
| ElectricityAccessCode | H4 | Binary factor | Has/doesn't-have access. |
| SocialGrantRecipients | H5 | Numeric count, plus a derived binary flag | Count may show a dose-response effect a binary flag would miss. |
| MainToiletCode | H6 | Categorical factor; consider collapsing to improved/unimproved | Matches H6's framing as a poverty proxy. |

## 4.6 Train/Test Split Strategy

A stratified 70/30 split on `Vulnerability_Tier` is used, rather than a
plain random split, so the minority tiers are proportionally represented
in both sets. The split runs on the 20,925 households with a valid FI
Score -- the 15 unscorable households are excluded before splitting, so
they cannot appear in either set.

```{r split}
set.seed(381)
split <- initial_split(modelling_data, prop = 0.7, strata = Vulnerability_Tier)
train <- training(split)
test  <- testing(split)

cat("Train:", nrow(train), " Test:", nrow(test), "\n")
cat("\nTrain tier proportions (%):\n")
print(round(prop.table(table(train$Vulnerability_Tier)) * 100, 2))
cat("\nTest tier proportions (%):\n")
print(round(prop.table(table(test$Vulnerability_Tier)) * 100, 2))
```

## 4.7 Before/After Validation Evidence

```{r validation}
cat("Row count accounted for at every step:\n")
cat("  Integrated households:", nrow(integrated), "\n")
cat("  Excluded as unscorable:", n_unscorable, "\n")
cat("  Retained for modelling:", nrow(modelling_data), "\n")
stopifnot(nrow(train) + nrow(test) == nrow(modelling_data))
cat("  Train + test == retained:", nrow(train) + nrow(test) == nrow(modelling_data), "\n")
```

- Row count accounted for at every step: 20,940 integrated households
  in, 15 excluded as unscorable and logged (Section 4.1a), 20,925
  retained for target construction and the train/test split.
- FI Score distribution on the 20,925 retained households matches the
  Milestone 1 raw-file report within a small margin, confirming NA-aware
  scoring is consistent with the earlier raw-file scoring.
- Income summary before/after cleaning shows the extreme right skew
  removed by the log transform (median well below mean pre-transform).
- Class balance confirmed numerically for both target framings (Section
  4.2/4.3).
- Train/test tier proportions checked against the population split
  above, confirming stratification held.

## Export

```{r export}
final_cols <- c("HouseholdID", "FI_Score", "Vulnerability_Tier", "Food_Insecure_Binary",
                predictor_vars, "Income_Log", "Income_PerCapita")
stopifnot(all(final_cols %in% names(modelling_data)))

saveRDS(modelling_data[, final_cols], file.path(OUT_DIR, "M2_final_modelling_dataset.rds"))
write_csv(modelling_data[, final_cols], file.path(OUT_DIR, "M2_final_modelling_dataset.csv"))
saveRDS(train, file.path(OUT_DIR, "M2_train.rds"))
saveRDS(test,  file.path(OUT_DIR, "M2_test.rds"))

cat("Exported M2_final_modelling_dataset.rds/csv, M2_train.rds, M2_test.rds to", OUT_DIR, "\n")
```

## Data dictionary: derived variables

The data dictionary written by Person B's notebook documents the 122 integrated columns. The six
variables derived in this notebook (`FI_Score`, `n_fi_answered`, `Vulnerability_Tier`,
`Food_Insecure_Binary`, `Income_Log`, `Income_PerCapita`) were documented by hand in Milestone 2.
Those rows are kept in `data/reference/M2_derived_variables_dictionary.csv` and appended here, so
re-running Milestone 2 rebuilds `BIN381_M2_Data_Dictionary.csv` exactly as it was submitted.

```{r dictionary-derived}
dict_file  <- file.path(OUT_DIR, "BIN381_M2_Data_Dictionary.csv")
dictionary <- read_csv(dict_file, na = character(), show_col_types = FALSE)
derived    <- read_csv("data/reference/M2_derived_variables_dictionary.csv", na = character(), show_col_types = FALSE)
if (!any(derived$Variable %in% dictionary$Variable)) {
  write_excel_csv(bind_rows(dictionary, derived), dict_file)
}
cat("Data dictionary:", nrow(read_csv(dict_file, na = character(), show_col_types = FALSE)), "variables\n")
```
