---
title: "EDA"
output: html_document
---

Load data into usable data frames.
We did this for the following data sets D04 Dwelling Conditions, D05 Basic Services, D06 Household Energy, D07 Household Well being, D09 Household Resources and D10 Mobility Access 

```{r}

library(dplyr)

D04 <- read.csv("data/D04_Dwelling_Conditions.csv")
D05 <- read.csv("data/D05_Basic_Services.csv")
D06 <- read.csv("data/D06_Household_Energy.csv")
D07 <- read.csv("data/D07_Household_Wellbeing.csv")
D09 <- read.csv("data/D09_Household_Resources.csv")
D10 <- read.csv("data/D10_Mobility_Access.csv")

#save all the files into a list so that i can iterate through all of them in for loops later on 
datafiles <- list(D04=D04,D05=D05,D06=D06,D07=D07, D09=D09,D10=D10)


for (name in names(datafiles)) {
  file <- datafiles[[name]]
  cat("\n========== ", name, " ==========\n", sep = "")

  cat("\n-- str --\n");        str(file)          # str prints itself
  cat("\n-- dim --\n");        print(dim(file))   # everything else needs print() as otherwise they would just execute silently in the background and not show anythin 
  cat("\n-- names --\n");      print(names(file))
  cat("\n-- head --\n");       print(head(file))
  cat("\n-- tail --\n");       print(tail(file))
  cat("\n")
  
 
}
```
# Summary of all Data Set points
#### this would be easier to see all grouped together so therefore we use the summary function
```{r}
for (name in names(datafiles)) {
  file <- datafiles[[name]]
   cat("\n-- Summary of numbers --\n");  print(summary(file))
}
```
# Now for the Standard Deviations for each dataset points
#### this is split up per column if possible, code is explain in the code block
```{r}
cat("Standard Deviation for each dataset")
for (name in names(datafiles)) {
  file <- datafiles[[name]]
  # now i am going to make a function print, that will only print the numeric value columns standard deviations, if i dont do this, then i will get a run time error as some of the columns are characters and not numeric, and standard deviation cant be done on character values.
  print(sapply(file, function(col) if (is.numeric(col)) sd(col, na.rm = TRUE) else NA))
}
```
# Now for the range of all columns for the respective data sets
```{r}
cat("The range for each data set :")
for (name in names(datafiles)) {
  file <- datafiles[[name]]
  #same applies here as above, just now we are performing the quantile function on all the numeric values
  print(sapply(file, function(col) if (is.numeric(col)) quantile(col, na.rm = TRUE) else NA))
}
```
# Categorical Counts and Levels
```{r}
# For each categorical column, show the count AND its proportion (% share) side by side
for (name in names(datafiles)) {
  file <- datafiles[[name]]
  cat("\n====", name, "====\n")
  for (colName in names(file)) {
     if (length(unique(file[[colName]])) <= 20) {   # only true categorical columns (skip IDs / high-cardinality)
        cat("\n--", colName, "--\n")
        tbl  <- table(file[[colName]], useNA = "always")   # raw counts (NA counted too)
        prop <- round(prop.table(tbl), 3)                  # proportion of the total, 3 decimals
        # stack the two rows into one table so each count sits above its proportion
        print(rbind(count = tbl, prop = prop))
     }
  }
}
```
# Missing Values
```{r}
for (name in names(datafiles)) {
  file <- datafiles[[name]]
  cat("\n====", name, "====\n")
  print(colSums(is.na(file)))       # NA count for every column at once (no inner loop needed)
  cat("Rows with any NA:", sum(!complete.cases(file)), "\n")  # rows missing at least one value
}

```
#Graphing
##Histograms and Box Plots
```{r}

hist(D07$TotalMonthlyHouseholdIncomeRaw)
hist(D09$TotalMonthlyHouseholdIncomeRaw)

hist(D09$RequiredIncomeRaw)

hist(D04$RentRaw)

# TimeToWaterRaw is stored as CHARACTER because it contains "N/A" / "not recorded".
# as.numeric() coerces it; the text values become NA (you'll see a coercion warning). DATA-QUALITY FINDING.
hist(as.numeric(D05$TimeToWaterRaw), main = "Time to Water", xlab = "Minutes")
hist(D10$TimeToHealthFacilityRaw)
hist(D10$TaxiCostRaw)
hist(D10$TaxiDistanceRaw)

# formula interface: write the column name only (NOT D07$...) when using data =
# outline = FALSE hides extreme outliers so the boxes stay readable (income is very skewed)
# las = 2 turns the province labels vertical so all 9 fit
boxplot(TotalMonthlyHouseholdIncomeRaw ~ Province, data = D07, outline = FALSE, las = 2)
boxplot(TotalMonthlyHouseholdIncomeRaw ~ Province, data = D09, outline = FALSE, las = 2)

boxplot(D09$RequiredIncomeRaw)

boxplot(D04$RentRaw)

boxplot(as.numeric(D05$TimeToWaterRaw), main = "Time to Water")  # character col -> numeric (text becomes NA)
boxplot(D10$TimeToHealthFacilityRaw)
boxplot(D10$TaxiCostRaw)
boxplot(D10$TaxiDistanceRaw)
```
##Discrete Counts
```{r}
# HeadAge has a wide range, so a histogram (binned) reads best
hist(D04$HeadAge, main = "Head-of-Household Age", xlab = "Age (years)", col = "steelblue")

# The rest are small-range counts (0,1,2,3...), so a barplot of table() is clearer than a histogram
barplot(table(D04$HouseholdSize),            main = "Household Size",        xlab = "People",    ylab = "Households", col = "steelblue")
barplot(table(D04$TotalRooms),               main = "Rooms per Dwelling",    xlab = "Rooms",     ylab = "Households", col = "steelblue")
barplot(table(D09$TotalVehicles),            main = "Vehicles per Household",xlab = "Vehicles",  ylab = "Households", col = "steelblue")
barplot(table(D09$SalaryEarnersInHousehold), main = "Salary Earners",        xlab = "Earners",   ylab = "Households", col = "steelblue")
barplot(table(D09$SocialGrantRecipients),    main = "Social Grant Recipients",xlab = "Recipients",ylab = "Households", col = "steelblue")
```
##Categorical Bar Charts
```{r}
# barplot() needs a table() first — it plots the counts of each category, not the raw column.
# las = 2 keeps axis labels readable when there are several categories.

# --- Food security (D07) ---
barplot(table(D07$AdultHungerCode),            main = "Adult Hunger",                 xlab = "Code", ylab = "Households", col = "tomato")
barplot(table(D07$ChildHungerCode),            main = "Child Hunger",                 xlab = "Code", ylab = "Households", col = "tomato")
barplot(table(D07$WorriedFoodWouldRunOutCode), main = "Worried Food Would Run Out",   xlab = "Code", ylab = "Households", col = "tomato")

# --- Services (D05, D06) ---
barplot(table(D06$ElectricityAccessCode),      main = "Electricity Access",           xlab = "Code", ylab = "Households", col = "darkorange")
barplot(table(D06$LoadsheddingFrequencyCode),  main = "Loadshedding Frequency",       xlab = "Code", ylab = "Households", col = "darkorange")
barplot(table(D05$DrinkingWaterSourceCode),    main = "Drinking Water Source",        xlab = "Code", ylab = "Households", col = "darkorange", las = 2)
barplot(table(D05$WaterSafeCode),              main = "Water Considered Safe",        xlab = "Code", ylab = "Households", col = "darkorange")
barplot(table(D05$MainToiletCode),             main = "Main Toilet Type",             xlab = "Code", ylab = "Households", col = "darkorange", las = 2)

# --- Economic (D04, D09) ---
barplot(table(D09$MainIncomeSourceCode),       main = "Main Income Source",           xlab = "Code", ylab = "Households", col = "seagreen", las = 2)
barplot(table(D09$ExpenditureBandCode),        main = "Expenditure Band",             xlab = "Code", ylab = "Households", col = "seagreen")
barplot(table(D04$TenureCode),                 main = "Tenure Type",                  xlab = "Code", ylab = "Households", col = "seagreen")
```





















